---
title: Intraspecific variation in the rate of acclimation in the widespread copepod *Skistodiaptomus pallidus*
date: "`r Sys.Date()`"
output: 
  html_document:
          code_folding: hide
          code_download: true
          toc: true
          toc_float: true
  github_document:
          html_preview: false
          toc: true
          toc_depth: 3
---

```{r setup, include=T, message = F, warning = F, echo = F}
knitr::opts_chunk$set(
  echo = knitr::is_html_output(),
  fig.align = "center",
  fig.path = "../Figures/markdown/",
  dev = c("png", "pdf"),
  message = FALSE,
  warning = FALSE,
  collapse = T
)

theme_matt = function(base_size = 18,
                      dark_text = "grey20"){
  mid_text =  monochromeR::generate_palette(dark_text, "go_lighter", n_colours = 5)[2]
  light_text =  monochromeR::generate_palette(dark_text, "go_lighter", n_colours = 5)[3]
  
  ggpubr::theme_pubr(base_family="sans") %+replace% 
    theme(
      panel.background  = element_rect(fill="transparent", colour=NA), 
      plot.background = element_rect(fill="transparent", colour=NA), 
      legend.background = element_rect(fill="transparent", colour=NA),
      legend.key = element_rect(fill="transparent", colour=NA),
      text = element_text(colour = mid_text, lineheight = 1.1),
      title = element_text(size = base_size * 1.5,
                           colour = dark_text),
      axis.text = element_text(size = base_size,
                               colour = mid_text),
      axis.title.x = element_text(size = base_size * 1.2,
                                  margin = unit(c(3, 0, 0, 0), "mm")),
      axis.title.y = element_text(size = base_size * 1.2,
                                  margin = unit(c(0, 5, 0, 0), "mm"), 
                                  angle = 90),
      legend.text = element_text(size=base_size * 0.9),
      legend.title = element_text(size = base_size * 0.9, 
                                  face = "bold"),
      plot.margin = margin(0.25, 0.25, 0.25, 0.25,"cm")
    )
}

theme_matt_facets = function(base_size = 18,
                             dark_text = "grey20"){
  mid_text =  monochromeR::generate_palette(dark_text, "go_lighter", n_colours = 5)[2]
  light_text =  monochromeR::generate_palette(dark_text, "go_lighter", n_colours = 5)[3]
  
  theme_bw(base_family="sans") %+replace% 
    theme(
      panel.grid = element_blank(),
      panel.background  = element_rect(fill="transparent", colour=NA), 
      plot.background = element_rect(fill="transparent", colour=NA), 
      legend.background = element_rect(fill="transparent", colour=NA),
      legend.key = element_rect(fill="transparent", colour=NA),
      text = element_text(colour = mid_text, lineheight = 1.1),
      strip.text.x = element_text(size = base_size),
      title = element_text(size = base_size * 1.5,
                           colour = dark_text),
      axis.text = element_text(size = base_size,
                               colour = mid_text),
      axis.title.x = element_text(size = base_size * 1.2,
                                  margin = unit(c(3, 0, 0, 0), "mm")),
      axis.title.y = element_text(size = base_size * 1.2,
                                  margin = unit(c(0, 5, 0, 0), "mm"), 
                                  angle = 90),
      legend.text = element_text(size=base_size * 0.9),
      legend.title = element_text(size = base_size * 0.9, 
                                  face = "bold"),
      plot.margin = margin(0.25, 0.25, 0.25, 0.25,"cm")
    )
}
```

## Background

This project examines the rate and magnitude of acclimation in two populations of *Skistodiaptomus pallidus*.

## Results

### Incubator Temperatures

```{r acc-temps}
acc_windows = ctmax_data %>% 
  group_by(exp_rep) %>% 
  summarise(start_time = min(datetime), 
            end_time = max(datetime)) 

filtered_temps = inc_temps %>% 
  mutate(exp_rep = if_else(exp_rep == 0, 0.5, exp_rep)) %>% 
  left_join(acc_windows, by = "exp_rep") %>% 
  filter(datetime > start_time & datetime < end_time) %>% 
  mutate("treatment" = if_else(incubator_temp == 16, "control", "warming")) %>% 
  select(exp_rep:temp_c, treatment)

exp_temps = filtered_temps %>%  
  group_by(exp_rep, treatment) %>% 
  summarise(mean_temp = mean(temp_c), 
            temp_sd = sd(temp_c)) %>% 
  select(exp_rep, treatment, mean_temp, temp_sd) |> 
  bind_rows(data.frame(exp_rep = 5, 
                       treatment = "control", ### Manually adding failed temperature log 
                       mean_temp = 16.3, ## Temperature recorded at the final timepoint
                       temp_sd = NA))

exp_temps %>% 
  pivot_wider(id_cols = c(exp_rep), 
              names_from = treatment, 
              values_from = mean_temp) %>%
  mutate(temp_diff = warming - control) %>% 
  knitr::kable(digits = 2)
```

```{r temp-profiles}

ggplot(filtered_temps, aes(x = datetime, y = temp_c, color = factor(incubator_temp))) + 
  facet_wrap(exp_rep~., scales = "free_x") + 
  geom_hline(yintercept = c(16, 22)) + 
  geom_line(linewidth = 2) + 
  scale_colour_manual(values = c("royalblue", "brown2")) + 
  labs(x = "Date", 
       y = "Temperature (°C)", 
       colour = "Incubator Set Temp.") + 
  theme_matt_facets() + 
  theme(legend.position = "bottom")

```

### Data Summary

All experimental data is shown below.

```{r all-data-reps, fig.width=12, fig.height=5.3}
ctmax_data %>% 
  mutate(acc_hours = acc_hours + 0.1) %>% 
  ggplot(aes(x = acc_hours, y = ctmax, colour = treatment)) + 
  facet_grid(pop~exp_rep) +
  geom_point(size = 2) + 
  geom_smooth(method = "lm", formula = y ~ log(x)) + 
  scale_colour_manual(values = c("control" = "royalblue",
                                 "warming" = "brown2")) + 
  labs(x = "Acclimation Hour", 
       y = "CTmax (°C)") + 
  scale_x_continuous(breaks = c(0,100, 200)) + 
  theme_matt_facets() + 
  theme(legend.position = "bottom")
```

The experimental replicates are combined here.

```{r all-data-combined, fig.width=4, fig.height=5}

ctmax_data %>% 
  mutate(acc_hours = acc_hours + 0.1) %>% 
  ggplot(aes(x = acc_hours, y = ctmax, colour = treatment)) + 
  facet_grid(pop~.) +
  geom_point(size = 2) + 
  geom_smooth(method = "lm", formula = y ~ log(x)) + 
  scale_colour_manual(values = c("control" = "royalblue",
                                 "warming" = "brown2")) + 
  labs(x = "Acclimation Hour", 
       y = "CTmax (°C)") + 
  theme_matt_facets() + 
  theme(legend.position = "bottom")

```

The plot above shows the raw data. The plot below shows daily averages for each replicate.

```{r rep-averages, fig.width=6, fig.height=8}

ctmax_data %>% 
  mutate(acc_hours = acc_hours + 0.1) %>% 
  group_by(pop, treatment, exp_rep, acc_hours, acc_day) %>% 
  summarise(rep_ctmax = mean(ctmax)) %>% 
  ggplot(aes(x = acc_hours, y = rep_ctmax, colour = treatment)) + 
  facet_grid(pop~.) +
  geom_point(size = 2) + 
  geom_smooth(method = "lm", formula = y ~ log(x)) + 
  scale_colour_manual(values = c("control" = "royalblue",
                                 "warming" = "brown2")) + 
  labs(x = "Acclimation Hour", 
       y = "CTmax (°C)") + 
  theme_matt_facets() + 
  theme(legend.position = "bottom")

```

The plot below shows the overall average (average of the averages). Error bars represent the standard error (calculated on the daily mean CTmax values for the experimental replicates). Note, standard errors cannot be calculated for days 5, 6, and 8, as CTmax was only measured on these days during the initial replicate.

```{r exp-averages, fig.width=6, fig.height=8}

ctmax_data %>% 
  mutate(acc_hours = acc_hours + 0.1,
         acc_day = acc_day + 0.1) %>% 
  group_by(pop, treatment, exp_rep, acc_hours, acc_day) %>% 
  summarise(rep_ctmax = mean(ctmax)) %>% 
  ungroup() |> 
  group_by(pop, treatment, acc_day) |> 
  summarise(exp_ctmax = mean(rep_ctmax), 
            ctmax_se = sd(rep_ctmax)/ sqrt(n())) %>%
  ggplot(aes(x = acc_day, y = exp_ctmax, colour = treatment)) + 
  facet_grid(pop~.) +
    geom_line(stat = "smooth", method = "lm", formula = y ~ log(x), 
              alpha = 0.3, linewidth = 3) + 
  geom_point(size = 2) + 
  geom_errorbar(aes(ymin = exp_ctmax - ctmax_se, ymax = exp_ctmax + ctmax_se),
                width = 0.2) + 
  scale_colour_manual(values = c("control" = "royalblue",
                                 "warming" = "brown2")) + 
  labs(x = "Acclimation Hour", 
       y = "CTmax (°C)") + 
  theme_matt_facets() + 
  theme(legend.position = "bottom")

```

Shown below are the CTmax reaction norms for the two populations, grouped by acclimation day and experimental replicate.

```{r reaction-norms}

ctmax_data %>% 
  group_by(pop, treatment, acc_day, exp_rep) %>% 
  summarise(mean_ctmax = mean(ctmax, na.rm = T), 
            ctmax_se = sd(ctmax, na.rm = T)/sqrt(n())) %>% 
  mutate(group_id = paste(pop, exp_rep, acc_day)) %>% 
  ggplot(aes(x = treatment, y = mean_ctmax, colour = acc_day, group = group_id)) + 
  facet_grid(.~pop) + 
  geom_point(size = 2) + 
  geom_line(linewidth = 1.5) + 
  viridis::scale_color_viridis(option = "G", direction = -1) + 
  theme_matt_facets()


```

### Linear Models and Contrasts

We will be using a linear model to analyze the data: CTmax as a function of treatment, population, and acclimation day (with all possible interactions). We've also included random intercepts for the experimental replicates and tube number (as a proxy for position in the water bath).

```{r lmer}

model_data = ctmax_data %>%
  mutate(acc_hours = acc_hours + 0.1,
         acc_day = acc_day + 0.1, 
         exp_rep = as.factor(exp_rep)) 

mixed.model = lmer(ctmax ~ log(acc_day) * treatment * pop + 
                     (1 | exp_rep) + (1 | tube), 
                   data = model_data)
```

This model performs well.

```{r model-performance, fig.height=11, fig.width=9}
performance::check_model(mixed.model)
```

The model indicates a significant effect of treatment and population. A significant two way interaction between acclimation time and treatment indicates that CTmax in the control and warming treatments change in different ways (as expected). The significant three-way interaction between acclimation time, treatment, and population indicates that the different populations exhibit different acclimation dynamics (rate or magnitude of acclimation).

```{r ANOVA}
#summary(mixed.model)

car::Anova(mixed.model, type = "III") 
```

To explore the significant three way interaction, we used the linear model to calculate estimated marginal means for each treatment-population combination on each day. We then calculate contrasts between the control and warming treatment groups for each day. A positive contrast indicates an increase in CTmax in the warming group relative to the control group (i.e. an increase in thermal limits after acclimation to higher temperatures) on that day.

These contrasts provide one potential approach for estimating the rate and magnitude of acclimation capacities using the mathematical framework proposed by Burton and Einum.

Below is a rough approximate of what that might look like, with a logarithmic relationship shown between the effect of acclimation and the acclimation duration.

```{r rough-contrast-plot, fig.width=10, fig.height=5}

cat_model_data = ctmax_data %>%
  mutate(acc_hours = as.factor(acc_hours), 
         acc_day = as.factor(acc_day), 
         exp_rep = as.factor(exp_rep))
  

cat_mixed.model = lmer(ctmax ~ acc_day * treatment * pop + 
                         (1 | exp_rep) + (1 | tube), 
                       data = cat_model_data)


contrasts = emmeans::emmeans(cat_mixed.model, ~ treatment | pop * acc_day) %>% 
  emmeans::contrast("revpairwise") %>% as_tibble() %>% 
  mutate(acc_day = as.numeric(as.character(acc_day)))

#write.csv(contrasts, file = "Output/Output_data/contrasts.csv", row.names = F)

contrasts %>% 
  mutate(acc_day = if_else(acc_day == 0, 0.01, acc_day)) %>% 
  ggplot(aes(x = acc_day, y = estimate)) + 
  facet_wrap(pop~.) + 
  geom_hline(yintercept = 0) +
  geom_errorbar(aes(ymin = estimate - SE, ymax = estimate + SE), 
                linewidth = 1, width = 0.3) + 
  geom_point(size = 3) + 
  geom_smooth(method = "lm", formula = y ~ log(x)) + 
  labs(x = "Acc. Day", 
       y = "Contrast (Warming - Control; °C)") + 
  theme_matt_facets()
```

### Parameter estimation

Burton and Einum (2025) describe an approach for measuring both rate of acclimation and acclimation capacity from a time series of CTmax values.

Their approach relies on fitting the following model to the data: $Z_t = Z_a * (1-e^{−λt})$

In this model, $Z_t$ is the effect of acclimation at time $t$. $Z_a$ is the fully acclimated CTmax value (the asymptotic value, i.e. the parameter representing magnitude of acclimation). The parameter $λ$ is the rate of acclimation. In the Burton and Einum study, data sets had to be transformed such that measurements at the different time points were all relative to the first CTmax measurements (e.g. all time series start at zero and measure the change relative to the start point).

We will take two approaches here: 1) estimating these parameters for each experimental replicate (warming treatment only) to provide an average rate and magnitude for each population, and 2) estimating these parameters using the estimated contrasts from the linear mixed effects model.

#### Approach 1 - Raw Data

```{r raw-data-params}
# Zt = Za * (1-e^(−λt))
# Za is the rescaled asymptotic critical temperature when acclimation is complete (i.e., plasticity capacity)
# λ is the plasticity rate (per hour)

raw_param_data = ctmax_data %>%
  filter(treatment == "warming" & exp_rep != 0.5) %>% 
  mutate(acc_hours = if_else(acc_hours == 0, 0.01, acc_hours)) 

rep_params = data.frame()
rep_means = data.frame()

for (i in 1:length(unique(raw_param_data$pop))){
  
  pop_data = filter(raw_param_data, pop == unique(raw_param_data$pop)[i]) %>% 
    drop_na() 
  
  for(rep in unique(pop_data$exp_rep)){
    
    rep_control = filter(exp_temps, exp_rep == rep, treatment == "control")
    rep_warming = filter(exp_temps, exp_rep == rep, treatment == "warming")
    
    rep_data = filter(pop_data, exp_rep == rep) %>% 
      group_by(exp_rep, pop, acc_hours) %>%  
      summarise(ctmax = mean(ctmax)) %>% 
      ungroup() %>% 
      mutate(adj_ctmax = ctmax - first(ctmax)) %>% 
      filter(adj_ctmax >= 0)
    
    rep_means = bind_rows(rep_means, rep_data)
    
    mod1 = try(nls.multstart::nls_multstart(
      adj_ctmax ~ z_asymp*(1-exp(-lambda*acc_hours)),
      data = rep_data,
      iter = 1000,
      start_lower = c(z_asymp=0.01, lambda=0),
      start_upper = c(z_asymp = 10, lambda=1),
      lower = c(z_asymp = 0, lambda=0),
      supp_errors = 'Y',
      convergence_count = FALSE,
      na.action = na.omit), silent =TRUE)
    
    
    fit_error = (is(mod1, 'try-error')|is(mod1,'error')) 
    
    if(fit_error==F){   #if model converged
      params = data.frame(
        pop = rep_data$pop[1],
        exp_rep = rep,
        num_contrasts = length(rep_data$adj_ctmax),
        z_asymp = summary(mod1)$coefficients[1,1],
        z_asymp_var = vcov(mod1)[1,1],
        lambda = summary(mod1)$coefficients[2,1],
        lambda.var = vcov(mod1)[2,2]) %>% 
        mutate(temp_diff = rep_warming$mean_temp-rep_control$mean_temp, 
               arr = z_asymp / (rep_warming$mean_temp-rep_control$mean_temp))
      
      rep_params = bind_rows(rep_params, params)
    }
    
  }
  
}

if(dim(rep_params)[1] > 0){
  rep_params %>% 
    select(pop, exp_rep, n = num_contrasts, temp_diff, z_asymp, arr, lambda) %>% 
    knitr::kable()
}

```

The plot here shows the estimated contrasts on each day. The model fit is included for both populations (in blue), along with the estimated final magnitude of acclimation (grey horizontal line). Note that in Rep 5, parameter estimates for the OP population were outside of reasonable estimates - the asymptote falls outside the plotted region.

```{r raw-data-param-plots, fig.width=12, fig.height=6}

rep_predictions = data.frame(
  pop = rep(rep_params$pop, each = 100), 
  exp_rep = rep(rep_params$exp_rep, each = 100), 
  z_asymp = rep(rep_params$z_asymp, each = 100), 
  lambda = rep(rep_params$lambda, each = 100),
  acc_hours = rep(seq(0, max(raw_param_data$acc_hours), length.out = 100)), times = dim(rep_params)[1]) %>% 
  ungroup() %>% 
  mutate(pred_ctmax = z_asymp * (1-exp(-lambda*acc_hours)))

rep_means %>% 
  filter(exp_rep != 0.5) %>% 
  ggplot(aes(x = acc_hours, y = adj_ctmax)) + 
  facet_grid(pop~exp_rep) + 
  geom_hline(yintercept = 0) +
  geom_hline(data = rep_params, aes(yintercept = z_asymp),
             colour = "grey") + 
  geom_point(size = 3) + 
  geom_line(data = rep_predictions, aes(x = acc_hours, y = pred_ctmax),
            colour = "blue",
            linewidth=1.5) + 
  coord_cartesian(ylim = c(-0.2, max(rep_means$adj_ctmax) + 0.3)) + 
  labs(x = "Acc. Hours", 
       y = "Change in CTmax (°C)") + 
  theme_matt_facets()

```

#### Approach 2 - Model Contrasts

The parameter estimates from the mixed effects model are shown below.

```{r model-params}
# Zt = Za * (1-e^(−λt))
# Za is the rescaled asymptotic critical temperature when acclimation is complete (i.e., plasticity capacity)
# λ is the plasticity rate (per hour)

param_data = contrasts %>%
  group_by(pop) %>% 
  arrange(acc_day) %>% 
  mutate(acc_day = if_else(acc_day == 0, 0.01, acc_day)) 

acc_params = data.frame()

for (i in 1:length(unique(param_data$pop))){
  
  pop_data = filter(param_data, pop == unique(param_data$pop)[i]) %>% 
    drop_na() 
  
  mod1 = try(nls.multstart::nls_multstart(
    estimate ~ z_asymp*(1-exp(-lambda*acc_day)),
    data = pop_data,
    iter = 1000,
    start_lower = c(z_asymp=0.01, lambda=0),
    start_upper = c(z_asymp = 10, lambda=1),
    lower = c(z_asymp = 0, lambda=0),
    supp_errors = 'Y',
    convergence_count = FALSE,
    na.action = na.omit), silent =TRUE)
  
  
  fit_error = (is(mod1, 'try-error')|is(mod1,'error')) 
  
  if(fit_error==F){   #if model converged
    pop_params = data.frame(
      z_asymp = summary(mod1)$coefficients[1,1],
      z_asymp_var = vcov(mod1)[1,1],
      lambda = summary(mod1)$coefficients[2,1],
      lambda.var = vcov(mod1)[2,2],
      pop = pop_data$pop[1],
      num_contrasts = length(pop_data$estimate)) %>% 
      mutate(arr = z_asymp / (22-16))
    
    acc_params = bind_rows(acc_params, pop_params)
  }
}

if(dim(acc_params)[1] > 0){
  acc_params %>% 
    select(pop, n = num_contrasts, z_asymp, arr, lambda) %>% 
    knitr::kable()
}

#write.csv(acc_params, file = "Output/Output_data/acc_params.csv", row.names = F)

```

The plot here shows the estimated contrasts on each day. The model fit is included for both populations (in blue), along with the estimated final magnitude of acclimation (grey horizontal line).

```{r model-param-plots, fig.width=10, fig.height=5}

if(dim(acc_params)[1] > 0){
  
  cp_params = filter(acc_params, pop == "CP")
  op_params = filter(acc_params, pop == "OP")
  
  # Create a new data frame with predicted values
  acc_day <- seq(0, max(param_data$acc_day), length.out = 100)
  cp_pred <- cp_params$z_asymp*(1-exp(-cp_params$lambda*acc_day))
  op_pred <- op_params$z_asymp*(1-exp(-op_params$lambda*acc_day))
  
  predictions = data.frame(acc_day, 
                           cp_pred, 
                           op_pred) %>% 
    pivot_longer(cols = c(cp_pred:op_pred), 
                 names_to = c("pop", NA), 
                 names_sep = "_", 
                 values_to = "pred") %>% 
    mutate(pop = toupper(pop))
  
  param_data %>% 
    ggplot(aes(x = acc_day, y = estimate)) + 
    facet_wrap(pop~.) + 
    geom_hline(yintercept = 0) +
    geom_hline(data = acc_params, aes(yintercept = z_asymp),
               colour = "grey") + 
    geom_errorbar(aes(ymin = estimate - SE, ymax = estimate + SE), 
                  linewidth = 1, width = 0.3) + 
    geom_point(size = 3) + 
    geom_line(data = predictions, aes(x = acc_day, y = pred),
              colour = "blue",
              linewidth=1.5)+
    labs(x = "Acc. Day", 
         y = "Contrast (Warming - Control; °C)") + 
    theme_matt_facets()
  
}
```

## Initial Conclusions

The three variables of interest are: CTmax (across all control replicates; tests for differences between populations); rate of acclimation (the lambda value); magnitude of acclimation (ARR values from the parameter estimates).

Two different approaches were used to estimate these parameters, using either the raw data adjusted in the manner proposed by Burton and Einum, or using the daily contrasts from a linear mixed effects model. The raw data approach failed to estimate reasonable parameters for one of the experimental replicates (due to the late timing of a small observed increase in CTmax in this replicate). Further, the raw data approach does not take into account changes in CTmax in the control treatment, which is particularly important given the relatively short generation times of these copepods - the 10 day acclimation period we use represents a significant portion of the generation time, and aging affects may therefore have an influence during the later period of an experiment. The linear mixed effects model accounts for this, along with across-replicate variation, making this the preferred method for estimating the parameters of interest.

## Still to do: 

- Additional populations

- CHELSA environmental data analysis

- Comparison between control data (population divergence)
