---
title: Rate of Acclimation in Skistodiaptomus pallidus
date: "`r Sys.Date()`"
output: 
  html_document:
          code_folding: hide
          code_download: true
          toc: true
          toc_float: true
  github_document:
          html_preview: false
          toc: true
          toc_depth: 3
---

```{r setup, include=T, message = F, warning = F, echo = F}
knitr::opts_chunk$set(
  echo = knitr::is_html_output(),
  fig.align = "center",
  fig.path = "../Figures/markdown/",
  dev = c("png", "pdf"),
  message = FALSE,
  warning = FALSE,
  collapse = T
)

theme_matt = function(base_size = 18,
                      dark_text = "grey20"){
  mid_text <-  monochromeR::generate_palette(dark_text, "go_lighter", n_colours = 5)[2]
  light_text <-  monochromeR::generate_palette(dark_text, "go_lighter", n_colours = 5)[3]
  
  ggpubr::theme_pubr(base_family="sans") %+replace% 
    theme(
      panel.background  = element_rect(fill="transparent", colour=NA), 
      plot.background = element_rect(fill="transparent", colour=NA), 
      legend.background = element_rect(fill="transparent", colour=NA),
      legend.key = element_rect(fill="transparent", colour=NA),
      text = element_text(colour = mid_text, lineheight = 1.1),
      title = element_text(size = base_size * 1.5,
                           colour = dark_text),
      axis.text = element_text(size = base_size,
                               colour = mid_text),
      axis.title.x = element_text(size = base_size * 1.2,
                                  margin = unit(c(3, 0, 0, 0), "mm")),
      axis.title.y = element_text(size = base_size * 1.2,
                                  margin = unit(c(0, 5, 0, 0), "mm"), 
                                  angle = 90),
      legend.text = element_text(size=base_size * 0.9),
      legend.title = element_text(size = base_size * 0.9, 
                                  face = "bold"),
      plot.margin = margin(0.25, 0.25, 0.25, 0.25,"cm")
    )
}

theme_matt_facets = function(base_size = 18,
                             dark_text = "grey20"){
  mid_text <-  monochromeR::generate_palette(dark_text, "go_lighter", n_colours = 5)[2]
  light_text <-  monochromeR::generate_palette(dark_text, "go_lighter", n_colours = 5)[3]
  
  theme_bw(base_family="sans") %+replace% 
    theme(
      panel.grid = element_blank(),
      panel.background  = element_rect(fill="transparent", colour=NA), 
      plot.background = element_rect(fill="transparent", colour=NA), 
      legend.background = element_rect(fill="transparent", colour=NA),
      legend.key = element_rect(fill="transparent", colour=NA),
      text = element_text(colour = mid_text, lineheight = 1.1),
      strip.text.x = element_text(size = base_size),
      title = element_text(size = base_size * 1.5,
                           colour = dark_text),
      axis.text = element_text(size = base_size,
                               colour = mid_text),
      axis.title.x = element_text(size = base_size * 1.2,
                                  margin = unit(c(3, 0, 0, 0), "mm")),
      axis.title.y = element_text(size = base_size * 1.2,
                                  margin = unit(c(0, 5, 0, 0), "mm"), 
                                  angle = 90),
      legend.text = element_text(size=base_size * 0.9),
      legend.title = element_text(size = base_size * 0.9, 
                                  face = "bold"),
      plot.margin = margin(0.25, 0.25, 0.25, 0.25,"cm")
    )
}
```

## Preliminary Trial 

The temperatures averages and standard deviations from the two incubators are shown in this table. 

```{r}
acc_windows = ctmax_data %>% 
  group_by(exp_rep) %>% 
  summarise(start_time = min(datetime), 
            end_time = max(datetime))

# To do: figure out a way to filter the data based on the acc_windows object; each exp_rep has a separate start and end time that should be used
inc_temps %>% 
  filter(datetime > acc_windows$start_time & datetime < acc_windows$end_time) %>%  
  group_by(incubator_temp) %>% 
  summarise(mean_temp = mean(temp_c), 
            temp_sd = sd(temp_c)) %>% 
  knitr::kable(digits = 2)
```

While the average temperature was closer to the intended temperature in the 22°C incubator, temperature was more variable than in the 16°C incubator. 

```{r fig.width=9, fig.height=6}

# Eventually will want to convert datetime into 'time since start' or something like that so all the reps can be overlaid
inc_temps %>% 
  filter(datetime > acc_windows$start_time & datetime < acc_windows$end_time) %>%  
  mutate(inc_id = paste(exp_rep, incubator_id, sep = "_")) %>% 
  ggplot(aes(x = datetime, y = temp_c, colour = factor(incubator_temp), group = inc_id)) + 
  geom_hline(yintercept = c(16,22)) + 
  geom_line(linewidth = 2) + 
  scale_colour_manual(values = c("royalblue", "brown2")) + 
  labs(y = "Temperature (°C)", 
       x = "Date", 
       colour = "Incubator \nTemp. (°C)") + 
  theme_bw(base_size = 25) + 
  theme(panel.grid = element_blank())
```

Thermal limit data is shown below. The larger points indicate the mean for each treatment group on each acclimation day, with the raw data shown as lighter points in the background. A general trend of increasing thermal limits in the warming acclimation group is present, but there is a fair amount of variation. That being said, it is too early to make any conclusions about specific patterns. 

```{r fig.width=10, fig.height=5.2}
ctmax_data %>% 
  group_by(pop, treatment, acc_day) %>% 
  summarise(mean_ctmax = mean(ctmax)) %>% 
ggplot(aes(x = acc_day, y = mean_ctmax, colour = treatment)) + 
  facet_wrap(pop~.) + 
  geom_point(data = ctmax_data, aes(y = ctmax),
             alpha = 0.3) + 
  geom_point(size = 3) + 
  geom_line(linewidth = 1.5) + 
  scale_colour_manual(values = c("control" = "royalblue",
                                 "warming" = "brown2")) + 
  labs(x = "Acclimation Day", 
       y = "CTmax (°C)") + 
  theme_matt_facets()
```

We will be using a linear model to analyze the data, and to examine specific patterns in the changes in thermal limits over time. For now, we use a simple linear regression model: CTmax as a function of treatment, population, and acclimation day (with all possible interactions). This model is limited by the small sample size, but performs reasonably well.

```{r fig.height=11, fig.width=9}
# this is an initial version of the model; later versions will include experimental replicate and incubator as random effects

model_data = ctmax_data %>% 
  mutate(acc_day = as.factor(acc_day))

prelim.model = lm(ctmax ~ acc_day * treatment * pop, data = model_data)

performance::check_model(prelim.model)
```

The model indicates a significant effect of acclimation time, treatment, and population, along with a significant interaction between acclimation time and treatment. 

```{r}
car::Anova(prelim.model) %>% knitr::kable()
```

We'll calculate estimated marginal means for each treatment group on each acclimation day. These marginal means can then be used to calculate pairwise contrasts between the control and warming groups (CTmax in the 22°C group - CTmax in the 16°C group). As such, a positive contrast indicates an increase in CTmax in the warming group relative to the control group. 

As for other analyses, it's still too early to draw any major conclusions from these results. It seems, however, like 6 days may not be enough for complete acclimation and that there may be differences between the two populations. 

```{r}
emmeans::emmeans(prelim.model, ~ treatment | pop * acc_day) %>% 
  emmeans::contrast("revpairwise") %>% as_tibble() %>% 
  mutate(acc_day = as.numeric(as.character(acc_day))) %>% 
  ggplot(aes(x = acc_day, y = estimate)) + 
  facet_wrap(pop~.) + 
  geom_hline(yintercept = 0) +
  geom_errorbar(aes(ymin = estimate - SE, ymax = estimate + SE), 
                linewidth = 1, width = 0.3) + 
  geom_point(size = 3) + 
  geom_smooth(method = "lm") + 
  labs(x = "Acc. Day", 
       y = "Contrast (Warming - Control; °C)") + 
  theme_matt_facets()

```



